---
title: 'Homework #3'
author: "Keene Morrow"
date: "5/8/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(janitor)

data <- read_csv("esm204_hw3_data.csv") %>%
  janitor::clean_names() %>%
  rename(q_low = q_low_gallons,
         q_high = q_high_gallons,
         price = price_dollars)
```

ESM 204: Incidence of Gas Tax

This homework asks you to examine the distributional effects of a gas tax. The overall questions are:

(1) What will be the distributional effects of the tax?

(2) How much revenue will be generated for infrastructure repairs?

(3) By how much will environmental damage be reduced and who will benefit?

(4) What is the role of electric vehicle technology?

You can use the following stylized facts:

- Consumers can be separated into two income groups: “High” and “Low”. The data set provides price (in $) and quantity (in gallons) estimates of demand per day for the two groups. Run regressions to estimate the demand curves for “High” and “Low” income consumers.
- Initially, there is no gas tax.
- The current gas price (without any taxes) is $3.00/gal.
- The marginal cost of producing a gallon of gasoline is linear and has price-intercept of 0. For the purposes of this exercise, you can assume that the country in question is self-contained in supply and demand (you don’t need to worry about global markets).
- Consuming each gallon of gasoline creates an environmental and health externality in the country of approximately \$1.50. In addition, it imposes an externality (from climate change) of about \$0.50 on the rest of the world.

```{r}
# Demand Regressions

#------------------
#Low Income Model
#------------------
low_lm <- lm(price ~ q_low, data = data)

print(low_lm)

# Sam's weird hard function for p
predict_low_price <- function(q, low_lm){
  b0 <- low_lm$coefficients["(Intercept)"]
  b1 <- low_lm$coefficients["q_low"]
  predicted_p_low <- b0 + b1 * q
  return(predicted_p_low)
}

# Sam's weird hard function for q
predict_low_q <- function(p, low_lm){
  b0 <- low_lm$coefficients["(Intercept)"]
  b1 <- low_lm$coefficients["q_low"]
  predicted_q_low <- (p - b0) / b1
  return(predicted_q_low)
}

# Easy function for p
low_p_fun <- function(q){
  predicted_p_low <- low_lm$coefficients["(Intercept)"] + low_lm$coefficients["q_low"] * q
  return(predicted_p_low)
}

# Easy function for q
low_q_fun <- function(p){
  predicted_q_low <- (p - low_lm$coefficients["(Intercept)"]) / low_lm$coefficients["q_low"]
  return(predicted_q_low)
}

# x intercept low income
xint_low_p <- (0 - low_lm$coefficients["(Intercept)"]) / low_lm$coefficients["q_low"]

#------------------
# High Income Model
#------------------
high_lm <- lm(price ~ q_high, data = data)

print(high_lm)

# Sam's weird hard function for p
predict_high_price <- function(q, high_lm){
  b0 <- high_lm$coefficients["(Intercept)"]
  b1 <- high_lm$coefficients["q_high"]
  predicted_p_high <- b0 + b1 * q
  return(predicted_p_high)
}

# Sam's weird hard function for q
predict_high_q <- function(p, high_lm){
  b0 <- high_lm$coefficients["(Intercept)"]
  b1 <- high_lm$coefficients["q_high"]
  predicted_q_high <- (p - b0) / b1
  return(predicted_q_high)
}

# Easy function for p
high_p_fun <- function(q){
  predicted_p_high <- high_lm$coefficients["(Intercept)"] + high_lm$coefficients["q_high"] * q
  return(predicted_p_high)
}

# Easy function for q
high_q_fun <- function(p){
  predicted_q_high <- (p - high_lm$coefficients["(Intercept)"]) / high_lm$coefficients["q_high"]
  return(predicted_q_high)
}

# x intercept high income
xint_high_p <- (0 - high_lm$coefficients["(Intercept)"]) / high_lm$coefficients["q_high"]

#------------------
# Predictions
#------------------
q_vector <- seq(0, 578800, by = 100)
p_vector <- seq(0, 16, by = 0.1)

predict_demand <- data.frame(q_vector) %>%
  mutate(low_price = predict_low_price(q_vector, low_lm),
         high_price =  predict_high_price(q_vector, high_lm)) %>%
  mutate(low_price = ifelse(low_price < 0, NA, low_price),
         high_price = ifelse(high_price < 0, NA, high_price))

predict_demand_p <- data.frame(p_vector) %>%
  mutate(low_q = predict_low_q(p_vector, low_lm),
         high_q =  predict_high_q(p_vector, high_lm))%>%
  mutate(low_q = ifelse(low_q < 0, NA, low_q),
         high_q = ifelse(high_q < 0, NA, high_q))

#------------------
# Status Quo Demand
#------------------
dmd_sq_low <- low_q_fun(3)
dmd_sq_high <- high_q_fun(3)

```

### Demand Equations

Low Income Demand:
MWTP_d = `r low_lm$coefficients["(Intercept)"]` + `r low_lm$coefficients["q_low"]` * q

High Income Demand:
MWTP_d = `r high_lm$coefficients["(Intercept)"]` + `r high_lm$coefficients["q_high"]` * q




Your assignment is to walk through the following analyses and then draft a memo to the Council of Economic Advisors advising them on the answers to the four questions posed at the top of this page. You should include a technical appendix with responses to the questions below:



1. What is the aggregate daily demand curve for gasoline? What is the supply curve for gasoline? What is the “benefit” to consumers under the status quo? What is the “benefit” to producers under the status quo? What is the environmental cost under the status quo (locally and in the rest of the world)?


```{r}
# Aggregate demand: sum horizontally
agg <- predict_demand_p %>%
  mutate(q_agg = ifelse(is.na(low_q), high_q, low_q + high_q))

# Easy function for q
agg_q_fun <- function(p){
  agg_q <- ((p - high_lm$coefficients["(Intercept)"]) / high_lm$coefficients["q_high"]) + ((p - low_lm$coefficients["(Intercept)"]) / low_lm$coefficients["q_low"])
  return(agg_q)
}

dmd_sq_agg_q <- agg_q_fun(3)

predict_agg_demand <- data.frame(p = p_vector) %>%
  mutate(q = agg_q_fun(p)) %>%
  filter(p <= 10)

agg_lm <- lm(p ~ q, data = predict_agg_demand)


agg_p_fun <- function(q){
  p <- agg_lm$coefficients["(Intercept)"] + agg_lm$coefficients["q"] * q
  return(p)
}

```

### Aggregate Demand Equation

q = ((p - `r high_lm$coefficients["(Intercept)"]` ) / `r high_lm$coefficients["q_high"]` ) + ((p - `r low_lm$coefficients["(Intercept)"]` ) / `r low_lm$coefficients["q_low"]` )

p = `r agg_lm$coefficients["(Intercept)"]` + `r agg_lm$coefficients["q"]` * q

```{r}
# Supply

#We need to draw a line between 0,0 and the aggregate demand intersection at $3.00. Then find the slope of that line.  That will give us the supply equation.

# also algebra looked good for aggregate and orignial p & q functions

supply_m <- 3 / dmd_sq_agg_q #slope

supply_fun <- function(q){
  p <- supply_m * q + 0
  return(p)
}

q_vector_sup <- seq(0, 750000, by = 100)

predict_supply <- data.frame(q = q_vector_sup) %>%
  mutate(p = supply_fun(q))

```

### Supply Equation
p = `r supply_m` * q


```{r}
# The Big Status Quo Plot
ggplot() +
  # low income demand
  geom_point(data = data,
             aes(x = q_low,
                 y = price),
             color = "blue") +
  geom_line(data = predict_demand,
            aes(x = q_vector,
                y = low_price),
            color = "blue") +
  geom_text(aes(x = 150000,
                y = 2.5),
            label = "Low",
            color = "blue",
            hjust = 0,
            angle = -55) +
  # high income demand
  geom_point(data = data,
             aes(x = q_high,
                 y = price),
             color = "dark green") +
  geom_line(data = predict_demand,
            aes(x = q_vector,
                y = high_price),
            color = "dark green") +
  geom_text(aes(x = 510000,
                y = 2.5),
            label = "High",
            color = "dark green",
            hjust = 0,
            angle = -32) +
  # aggregate demand
  geom_line(data = agg,
            aes(x = q_agg,
                y = p_vector),
            color = "black",
            linetype = "dashed") +
  geom_text(aes(x = 500000,
                y = 6),
            label = "Aggregate",
            hjust = 0,
            angle = -25) +
  geom_text(aes(x = 75000,
                y = 14.5),
            label = "Aggregate",
            hjust = 0,
            angle = -32) +
  # supply
  geom_line(data = predict_supply,
            aes(x = q,
                y = p),
            color = "purple") +
  geom_text(aes(x = 700000,
                y = 4),
            label = "Supply",
            color = "purple",
            hjust = 0,
            angle = 10) +
  # black line at $3.00/gallon for current gas price
  geom_hline(yintercept = 3,
             color = "grey50") +
  geom_text(aes(x = 0,
                y = 2.95),
            label = "Current Price",
            color = "grey50",
            hjust = 0,
            vjust = 1) +
  # Aesthetics
  labs(title = "Demand for Gasoline",
       y = "Price (dollars)",
       x = "Quantity (gallons)",
       caption = "\nESM 204 Spring 2020\nElliott Matthews & Keene Morrow") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal() +
  theme(plot.caption = element_text(hjust = 0, face = "italic"))
```




```{r}
# Benefit Plot

ggplot() +
  # aggregate demand
  geom_line(data = agg,
            aes(x = q_agg,
                y = p_vector),
            color = "black",
            linetype = "dashed") +
  geom_text(aes(x = 300000,
                y = 9.5),
            label = "Aggregate",
            hjust = 0,
            angle = -25) +
  geom_ribbon(data = agg %>% filter(q_agg < dmd_sq_agg_q),
              aes(x = q_agg,
                  ymax = p_vector,
                  ymin = 3),
              alpha = 0.5) +
  # supply
  geom_line(data = predict_supply,
            aes(x = q,
                y = p),
            color = "purple") +
  geom_text(aes(x = 700000,
                y = 4),
            label = "Supply",
            color = "purple",
            hjust = 0,
            angle = 10) +
  geom_ribbon(data = predict_supply %>% filter(q < dmd_sq_agg_q),
              aes(x = q,
                  ymax = 3,
                  ymin = p),
              alpha = 0.5,
              fill = "purple") +
  # black line at $3.00/gallon for current gas price
  geom_hline(yintercept = 3,
             color = "black") +
  # geom_text(aes(x = 0,
  #               y = 2.95),
  #           label = "Current Price",
  #           color = "black",
  #           hjust = 0,
  #           vjust = 1) +
  # Aesthetics
  labs(title = "Benefit",
       subtitle = "Status Quo",
       y = "Price (dollars)",
       x = "Quantity (gallons)",
       caption = "\nESM 204 Spring 2020\nElliott Matthews & Keene Morrow") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal() +
  theme(plot.caption = element_text(hjust = 0, face = "italic"))
```


```{r}
# Benefit to Consumers under Status Quo
benefit_c_sq <- (integrate(low_p_fun, 0, dmd_sq_low)$value - (dmd_sq_low * 3)) + (integrate(high_p_fun, 0, dmd_sq_high)$value - (dmd_sq_high * 3))
# I think this one is still good post supply considerations

# Benefit to Producers under Status Quo
benefit_p_sq <- integrate(supply_fun, 0, dmd_sq_agg_q)$value
#I think this needs to be altered too once the supply curve is calculated. It should be the triangle above the supply curve and below the market price. - GEM
# word - KM
# same as (dmd_sq_agg_q * 3)/2
# Yes, I agree ^ - GEM


# Environmental Cost under Status Quo

#Need to add the external cost listed in the problem statement to the supply curve we calculate. From there we solve for the new P and Q (the external cost shifts the supply curve upward, P goes up, Q goes down). The environmental cost is the rectangular area defined by PxQ. **** We do this for both local and world values **** - EM

# In other words, the environmental cost going to be the new supply function taking into account the environmental costs, integrated from 0 to the new intercept? - KM

#  Actually, my interpretation was whack. We don't add the externality onto the supply curve. The given externality will just be a flat curve and will represent the MEC.  Everything under the curve (0-1.50 vertically, and 0 - dm_sq_agg) will be there area we want. Sorry for the confusion- GEM

## Nationally

ntl_MEC <- 1.5

supply_fun_ntl_MEC <- function(q){
  p <- ntl_MEC
  return(p)
}

predict_supply_ntl_MEC <- data.frame(q = q_vector_sup) %>%
  mutate(p = supply_fun_ntl_MEC(q))

ntl_agg <- agg_q_fun(ntl_MEC) # intercept with aggregate demand

ntl_env_cost <- (1.5 - 0) * (ntl_agg - 0)

## Internationally

intl_MEC <- 0.5

supply_fun_intl_MEC <- function(q){
  p <- intl_MEC
  return(p)
}

predict_supply_intl_MEC <- data.frame(q = q_vector_sup) %>%
  mutate(p = supply_fun_intl_MEC(q))

intl_agg <- agg_q_fun(intl_MEC) # intercept with aggregate demand

intl_env_cost <- (1.5 - 0) * (intl_agg - 0)

# Plot
ggplot() +
  # aggregate demand
  geom_line(data = agg,
            aes(x = q_agg,
                y = p_vector),
            color = "black",
            linetype = "dashed") +
  geom_text(aes(x = 300000,
                y = 9.5),
            label = "Aggregate",
            hjust = 0,
            angle = -25) +
  # supply
  geom_line(data = predict_supply,
            aes(x = q,
                y = p),
            color = "purple") +
  geom_text(aes(x = 760000,
                y = 3.5),
            label = "Supply",
            color = "purple",
            hjust = 0,
            angle = 10) +
  geom_point(aes(x = dmd_sq_agg_q,
                 y = 3),
             color = "purple") +
  # supply with national cost
  geom_line(data = predict_supply_ntl_MEC,
            aes(x = q,
                y = p),
            color = "blue") +
  geom_text(aes(x = 760000,
                y = 1.5),
            label = "National",
            color = "blue",
            hjust = 0) +
  geom_point(aes(x = ntl_agg,
                 y = ntl_MEC),
             color = "blue") +
  # supply with international cost
  geom_line(data = predict_supply_intl_MEC,
            aes(x = q,
                y = p),
            color = "navy") +
  geom_text(aes(x = 760000,
                y = 0.5),
            label = "International",
            color = "navy",
            hjust = 0) +
  geom_point(aes(x = intl_agg,
                 y = intl_MEC),
             color = "navy") +
  # black line at $3.00/gallon for current gas price
  geom_hline(yintercept = 3,
             color = "black") +
  labs(title = "",
       subtitle = "",
       y = "Price (dollars)",
       x = "Quantity (gallons)",
       caption = "\nESM 204 Spring 2020\nElliott Matthews & Keene Morrow") +
  scale_x_continuous(expand = c(0, 150000)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal() +
  theme(plot.caption = element_text(hjust = 0, face = "italic"))

```

National Environmental Cost: `r ntl_env_cost`
Internationsl Environmental Cost: `r intl_env_cost`

2. How is the current consumer benefit divided between “High” and “Low” income consumers?

```{r}
benefit_low_sq <- (integrate(low_p_fun, 0, dmd_sq_low)$value - (dmd_sq_low * 3))
benefit_high_sq <- (integrate(high_p_fun, 0, dmd_sq_high)$value - (dmd_sq_high * 3))

pct_benefit_low_sq <- (benefit_low_sq / benefit_c_sq) * 100
pct_benefit_high_sq <- (benefit_high_sq / benefit_c_sq) * 100
```

Total Benefit: `r round(benefit_c_sq, 2)`

Low Income Benefit: `r round(benefit_low_sq, 2)`
Low Income Percent of Total Benefit: `r round(pct_benefit_low_sq, 2)`%

High Income Benefit: `r round(benefit_high_sq, 2)`
High Income Percent of Total Benefit: `r round(pct_benefit_high_sq, 2)`%


3. A gas tax of $1.00/gal. is proposed. What would be the effects of this tax on:

```{r}
ggplot() +
  # aggregate demand
  geom_line(data = agg,
            aes(x = q_agg,
                y = p_vector),
            color = "black",
            linetype = "dashed") +
  geom_text(aes(x = 300000,
                y = 9.5),
            label = "Aggregate",
            hjust = 0,
            angle = -25) +
  geom_ribbon(data = agg %>% filter(q_agg < 559090.9),
            aes(x = q_agg,
                ymax = p_vector,
                ymin = 3.794516),
            fill = "black",
            alpha = 0.5) +
  # supply
  geom_line(data = predict_supply,
            aes(x = q,
                y = p),
            color = "purple") +
  # geom_text(aes(x = 650000,
  #               y = 5.5),
  #           label = "Supply",
  #           color = "purple",
  #           hjust = 0,
  #           angle = 10) +
  geom_line(data = predict_supply,
            aes(x = q,
                y = p + 1),
            color = "red") +
  geom_ribbon(data = predict_supply %>% filter(q < 559090.9),
            aes(x = q,
                ymax = 3.794516 - 1,
                ymin = p),
            fill = "purple",
            alpha = 0.5) +
  # tax
  geom_ribbon(data = agg %>% filter(q_agg < 559090.9),
            aes(x = q_agg,
                ymax = 3.794516,
                ymin = 2.794516),
            fill = "red",
            alpha = 0.5) +
  
  # # black line at $3.00/gallon for current gas price
  geom_hline(yintercept = 3.794516,
             color = "black") +
  geom_hline(yintercept = 3.794516 - 1,
             color = "black") +
  geom_hline(yintercept = 3,
             color = "black") +
  geom_vline(xintercept = 559090.9,
             color = "black") +
  labs(title = "",
       subtitle = "",
       y = "Price (dollars)",
       x = "Quantity (gallons)",
       caption = "\nESM 204 Spring 2020\nElliott Matthews & Keene Morrow") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal() +
  theme(plot.caption = element_text(hjust = 0, face = "italic"))
```


a. The amount of gasoline produced and consumed.

```{r}
supply_tax_p_fun <- function(q){
  p <- supply_m * q + 1
  return(p)
}

supply_tax_q_fun <- function(p){
  q <- (p- 1) / supply_m
  return(q)
}

q_int_tax <- (agg_lm$coefficients["(Intercept)"] - 1)/(supply_m - agg_lm$coefficients["q"])
```

`r q_int_tax` gallons

b. The price of gasoline.

```{r}
p_int_tax <- agg_p_fun(q_int_tax)
supply_tax_p_fun(q_int_tax)
```

$`r round(p_int_tax, 2)` per gallon

c. Welfare of “High” income consumers.

```{r}
benefit_high_tax <- (integrate(high_p_fun, 0, 559090.9)$value - (559090.9 * )) # should be the 3.7 value
#revisit
```

`r benefit_high_tax`

d. Welfare of “Low” income consumers.

```{r}
benefit_low_tax <- (integrate(low_p_fun, 0, 559090.9)$value - (559090.9 * ))
#revisit
```

`r benefit_low_tax`

e. Welfare of gas producers.

```{r}
benefit_p_tax <- integrate(supply_fun, 0, 559090.9)$value
#revisit
```

`r benefit_p_tax`

f. Local environmental damage.

```{r}
## Nationally

ntl_MEC <- 1.5

supply_fun_ntl_MEC <- function(q){
  p <- ntl_MEC
  return(p)
}

predict_supply_ntl_MEC <- data.frame(q = q_vector_sup) %>%
  mutate(p = supply_fun_ntl_MEC(q))

ntl_agg <- agg_q_fun(ntl_MEC) # intercept with aggregate demand

ntl_env_cost <- (1.5 - 0) * (559090.9)


```


g. Rest of world environmental damage.

```{r}

## Internationally

intl_MEC <- 0.5

supply_fun_intl_MEC <- function(q){
  p <- intl_MEC
  return(p)
}

predict_supply_intl_MEC <- data.frame(q = q_vector_sup) %>%
  mutate(p = supply_fun_intl_MEC(q))

intl_agg <- agg_q_fun(intl_MEC) # intercept with aggregate demand

intl_env_cost <- (1.5 - 0) * (559090.9)

```


h. Total revenue generated by the tax.

```{r}

tax_revenue <- (3.7-2.7) * (559090.9)

```


4. Now, assume that all revenue from a tax will be redistributed to the two groups in proportion to their pre-tax consumption of gas. For example, if 80% of the gas was consumed by High income consumers, then they get 80% of the tax revenue. Also assume that “Low” income consumers bear all local environmental costs. For a range of gas taxes (ranging from \$0 - \$5.00/gal), calculate the effects of the tax on:

a. Overall welfare of “High” income consumers


b. Overall welfare of “Low” income consumers


c. Gas producers



5. A new electric car technology is invented and it lowers the demand curves of all income groups by half (vertically). Under these new demand curves, what are the effects on:

a. Gas consumption by “High” income consumers


b. Gas consumption by “Low” income consumers


c. Gas price


d. Local environmental damage from gasoline


e. Rest of world environmental damage from gasoline


6. Compare two situations: (1) Gas tax of $T/gal. but no electric car technology and (2) No gas tax but with electric car technology. What value of T (i.e. what gas tax) makes the local environmental quality equal between these two situations?

